Synthesis of Linear Functional Programs

### 2021-05-07 article/article.tex
more on background

\begin{enumerate}
\item General motivation
  \begin{itemize}
\item 
\item Specifications appear in many forms\dots
\item Program synthesis as a means of improving program
  correctness/programmer productivity.
\item Challenges: space of valid programs, space of valid
  specifications,  etc.
\end{itemize}
\item \emph{Type-based} program synthesis
  \begin{itemize}
 \item Rich types as specifications.
\item Languages with richer type systems allow for more precise
  types. This precision can statically eliminate various kinds of
  logical errors by making certain invalid program states ill-typed
  (e.g. a ``null aware'' type system, will ensure you cannot
  dereference a null-pointer).
\item But can also be a burden -- the whole point of these type
  systems is to ensure that ``less'' things are well-typed, and
  sometimes it's hard to convince the checker\dots
\item Type-based synthesis leverages rich types as a way of pruning
  the search space and by using types as a more ``familiar''
  specification.
\end{itemize}
\item ``My'' problem: Synthesis based on linear types
  \begin{itemize}
  \item Linear types constrain resource usage in programs by \emph{statically} limiting
    the number of times certain resources can be used during its
    lifetime.
  \item Applications in resource-aware programming such as concurrent
    programming (e.g. session types) and memory-management
    (e.g. Rust).
  \item This work will study program synthesis in a functional,
    linearly-typed setting, by leveraging the propositions-as-types
    correspondence between linear logic and the (linear)
    $\lambda$-calculus.
  \item What is props-as-types? What is linear logic? Linear $\lambda$
    is ``just'' a small functional core language.
  \item Through props-as-types, synthesis can be formulated as proof
    search and so we can leverage a lot of related technology from the
    literature.
  \end{itemize}
\item Goals:
  \begin{itemize}
\item Full and partial synthesis of well-typed (in the above sense)
  programs. Full synthesis means producing a function (or set of)
  satisfying the spec. Partial synthesis means being given a partial
  program (i.e. a function with a hole in it) and ``completing it''.
\item Start from a small system, build up to recursive
  types/functions.
\item Potential to go further: type refinements; interactive
  synthesis?
 \item Evaluation through expressiveness benchmarks (can we synthesize
   ``X''?); time measurements (how fast can we synthesize ``X''?');
   etc.
  \end{itemize}

\end{enumerate}

@@

>>>>

% \begin{enumerate}
% \item General motivation
%   \begin{itemize}
% \item 
% \item Specifications appear in many forms\dots
% \item Program synthesis as a means of improving program
%   correctness/programmer productivity.
% \item Challenges: space of valid programs, space of valid
%   specifications,  etc.
% \end{itemize}
% \item \emph{Type-based} program synthesis
%   \begin{itemize}
%  \item Rich types as specifications.
% \item Languages with richer type systems allow for more precise
%   types. This precision can statically eliminate various kinds of
%   logical errors by making certain invalid program states ill-typed
%   (e.g. a ``null aware'' type system, will ensure you cannot
%   dereference a null-pointer).
% \item But can also be a burden -- the whole point of these type
%   systems is to ensure that ``less'' things are well-typed, and
%   sometimes it's hard to convince the checker\dots
% \item Type-based synthesis leverages rich types as a way of pruning
%   the search space and by using types as a more ``familiar''
%   specification.
% \end{itemize}
% \item ``My'' problem: Synthesis based on linear types
%   \begin{itemize}
%   \item Linear types constrain resource usage in programs by \emph{statically} limiting
%     the number of times certain resources can be used during its
%     lifetime.
%   \item Applications in resource-aware programming such as concurrent
%     programming (e.g. session types) and memory-management
%     (e.g. Rust).
%   \item This work will study program synthesis in a functional,
%     linearly-typed setting, by leveraging the propositions-as-types
%     correspondence between linear logic and the (linear)
%     $\lambda$-calculus.
%   \item What is props-as-types? What is linear logic? Linear $\lambda$
%     is ``just'' a small functional core language.
%   \item Through props-as-types, synthesis can be formulated as proof
%     search and so we can leverage a lot of related technology from the
%     literature.
%   \end{itemize}
% \item Goals:
%   \begin{itemize}
% \item Full and partial synthesis of well-typed (in the above sense)
%   programs. Full synthesis means producing a function (or set of)
%   satisfying the spec. Partial synthesis means being given a partial
%   program (i.e. a function with a hole in it) and ``completing it''.
% \item Start from a small system, build up to recursive
%   types/functions.
% \item Potential to go further: type refinements; interactive
%   synthesis?
%  \item Evaluation through expressiveness benchmarks (can we synthesize
%    ``X''?); time measurements (how fast can we synthesize ``X''?');
%    etc.
%   \end{itemize}
% \end{enumerate}

@@

A type system can be formally described through natural deduction -- as a set of rules that determines the concept of deduction for some language~\cite{prawitznd65}.
This comes from the close relation between logic and programming languages, called the Curry-Howard isomorphism or correspondence. A logical proof describes a program, and a program describes a proof.
The main observations made are ``propositions as types'', ``proofs as programs'', ``proof checking as type checking'' \cite{https://www.cs.cmu.edu/~fp/courses/15312-f04/handouts/23-curryhoward.pdf}.
Formalizing our type system, the language i.e. the propositions of the logic are the types themselves, the inference rules are used to construct a proof which corresponds directly to a program, and the correctness of the proof assures the program is well-typed.
The typical natural deduction judgment
\[
    A \textrm{ is a proposition}
\]
is extended to include a \emph{term}
(i.e. to make proofs explicit with a formal definition -- these proofs map directly to a $\lambda$-calculus program as stated by the Curry-Howard correspondence????),
and modified to include a list of hypothesis. 
This extended judgment, called?? a typing judgment, has the form:
\[
    \Gamma \vdash M : A
\]
In the type system context, the list of hypothesis, $\Gamma$, is called \emph{typing environment} (i.e. set of variables and their types),
 the term $M$ is the \emph{program}, and $A$, the natural deduction proposition, is the \emph{type}.
 This judgment can be read -- $M$ has type $A$ in $\Gamma$.

Type rules assert the validity of certain judgments on the basis of other judgments already known to be valid \cite{http://homepage.cs.uiowa.edu/~tinelli/classes/185/Fall06/notes/cardelli-95.pdf}.
We introduce a few typing i.e. inference rules to display how a type system would be formalized.
In natural deduction, there are both \emph{introduction} and \emph{elimination} rules for almost all logical connectives.
The following rule is a hypothesis rule ??, não sei muito bem como sair daqui:)
\[
    \infer*[right=(var)]
    {  }
    {\Gamma, u: A \vdash u : A}
\]
It reads -- the program $u$ has type $A$ if our typing environment holds a variable labeled $u$ of type $A$.
[também não sei muito bem onde quero chegar com isto :)]
As another concrete example, let us describe the inference rule for the function type ($A \supset B$):
\[
    \infer*[right=($\supset I$)]
    {\Gamma, u: A \vdash M : B}
    {\Gamma \vdash \lambda u. M : A \supset B }
\]
It reads -- the program $\lambda u. M$ has type $A \supset B$ ($A$ implies $B$) as long as $M$ is a program of type $B$ knowning that $u$ is a variable of type $A$.
The rule shown above is an \emph{introduction} rule, that introduces the ``implication'' connective,

\section{Separate}
